GeckoGraph: A Visual Language for Polymorphic Types

### 2025-04-01 main.tex
Eval

%% 
%% Copyright 2007-2024 Elsevier Ltd
%% 
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%% 
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.3 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.3 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%% 
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%% 
%% Template article for Elsevier's document class `elsarticle'
%% with numbered style bibliographic references
%% SP 2008/03/01
%% $Id: elsarticle-template-num.tex 249 2024-04-06 10:51:24Z rishi $
%%
\documentclass[preprint,12pt]{elsarticle}

@@

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsmath package provides various useful equation environments.
\usepackage{amsmath}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.
\usepackage{lineno}

\journal{Journal of Computer Languages}

@@

%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for theassociated footnote;
%% use the fnref command within \author or \affiliation for footnotes;
%% use the fntext command for theassociated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for theassociated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
 \author{Shuai Fu}

@@

 \affiliation{organization={Monash University}}

%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \affiliation{organization={},
%%             addressline={},
%%             city={},
%%             postcode={},
%%             state={},
%%             country={}}
%% \fntext[label3]{}

@@

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{}
%% \affiliation[label1]{organization={},
%%             addressline={},
%%             city={},
%%             postcode={},
%%             state={},
%%             country={}}
%%
%% \affiliation[label2]{organization={},
%%             addressline={},
%%             city={},
%%             postcode={},
%%             state={},
%%             country={}}

@@

%% Add \usepackage{lineno} before \begin{document} and uncomment 
%% following line to enable line numbers
\linenumbers

@@

While type polymorphism is one of the oldest topics in programming language theory \cite{Cardelli1987-fp}, little research focuses on the \textit{usability} of polymorphic types. Hage argues that the expressiveness and power of type systems often come at the cost of usability~\cite{Hage2020-hg}. We aim to investigate the challenges of using polymorphic types and explore how to improve their usability with visualization and modern HCI techniques. To achieve this, we propose GeckoGraph, a graphical notation for types. GeckoGraph aims to complement traditional text-based type notation and make reading, understanding, and comparing types easier. GeckoGraph is prototyped and verified iteratively, leading to a design with visual clarity applicable to many programming contexts. Our study evaluating 714 participants' ability to solve type adaptation challenges in GeckoGraph versus text-based type annotation is, to our knowledge, the largest controlled user study of a functional programming tool or concept. The study results show that GeckoGraph helps improve programmers' ability to succeed in resolving the type challenges, especially for novice programmers.

@@

\subsection{Design of GeckoGraph}
The design of GeckoGraph focuses on visualizing types in functional languages (e.g., Haskell, ML). In this paper, we use Haskell as an example. As illustrated in this section, it can express basic types, polymorphic types, algebraic data types, and some advanced type-level features. However, GeckoGraph could also be used in imperative and multiparadigm languages such as Rust. We provide a GeckoGraph construction library for Haskell.

We identified three main design goals for GeckoGraph based on the challenges of using polymorphic types~\cite{Jun2000-ec, Jun2000-yu} and how programmers tend to use type annotations~\cite{Justin_Lubin2021-yy}, as follows.

\paragraph{\textbf{(D1) Low barrier to learn}}\label{goal1} GeckoGraph should take little to no effort to learn. The rules to translate a text-based type notation to GeckoGraph should be minimal. Where possible, GeckoGraph should be intuitive to programmers who are familiar with text-based type notation.

@@

\paragraph{\textbf{(D3) Easy to compare and search}} \label{goal3} GeckoGraph should aim to make the task of comparing two types easy, especially to make subtle differences in text-based notation harder to miss. This also includes the task of choosing an ideal function from a list of potential functions. For example, programmers search for a desired function from a documentation site with only partial knowledge of its type (e.g., the arity, one of the argument types, or type class it must fulfill).

\paragraph{Simple Types} 
Simple types, such as type variables and concrete types, are displayed in a cell \includegraphics[height=1em]{figures/SimpleType}: a solid-colored rectangular box with an angled corner on the top left. Each type identifier encodes a distinct color hue of the cell.  Its first 1 or 2 letters are displayed inside the cell at the bottom left to provide familiarity (design goal \ref{goal1}) and strong secondary encoding. The angled corner in the top left provides visual separation between two cells, even when the same color cells are next to each other, allowing GeckoGraph to be zoomed out to extremely small sizes (Section \ref{subsec:space}) without suffering readability (design goal \ref{goal2}).

@@

\begin{figure}[]
  \includegraphics[width=\linewidth]{figures/Indicator}
  \caption{
        \label{fig:indicator}
        An example of using the function indicator. The function indicator can be used to easily identify the arity of a function type by counting the connecting function indicators. For high-order functions where functions are arguments of other functions, it is very easy to see the ``order" of functions and how they are arranged. 
  }
\end{figure}

\paragraph{Type Classes} 
Type classes are an intrinsic part of Haskell \cite{Hudak2007-kn}, and many other functional languages. In GeckoGraph, the type classes (e.g., \texttt{ (A a, B a) => a}) are indicated in the extended area below one or more GeckoGraph cells \includegraphics[height=1.2em]{figures/TypeClass.png}. Each type class required on a type variable displays as a square indicator aligned on the right of the extended area. In GeckoGraph, type-class constraints are associated with every instance of the type variable that requires them. This means when displaying the type \texttt{(==) :: Eq a => a -> a -> Bool} in GeckoGraph, the constraint \texttt{Eq} appears in both occurrences of \texttt{a}.

The GeckoGraph type class's design promotes the type class placements rather than the type class names. Programmers can easily see where and how many type classes are required, but they may need an extra step (global legends of the color mapping or pop-up window) to identify the name of the type class. We believe that this decision is well justified. For example, when reading a type \texttt{(A a, A c, B a, B b, C b) => a -> b -> c}, programmers may need to switch back and forth to remember which type classes are needed on which variable. GeckoGraph helps minimize the effort to associate each type variable with all its type classes (design goal \ref{goal2}).

\subsection{Benefits of Using GeckoGraph}\label{sec:benefits}
\paragraph{Generalization patterns in type classes}
A frequently cited confusion among novices is the overlap between list type \texttt{[a]} and its more generic counterpart \texttt{Foldable} instances. The subtle differences are often not important for beginners. However, when encountering type errors in working with lists, Haskell often explains the error with the \texttt{Foldable} instance. For the example in Fig. \ref{fig:example-foldable} where a programmer mistakenly provided a \texttt{Char} instead of a \texttt{String} literal, the compiler shows an error message comparing the provided \texttt{Char} type with a confusingly named type \texttt{t0 a0}. Although any \texttt{Foldable} instance is perfectly suited for the list \cite{Waldmann2018-hu}, this generalization may reduce programmers' confidence in their understanding of the language and the ability to navigate out of a type error. In GeckoGraph, a list type and a type with a Foldable instance have the same shape. This allows the generic type \texttt{t0 a0} in the error message to assume the same shape as \texttt{[a]}. This generalization of concrete types and abstraction of type-class instances aims to allow for teaching fundamental functional programming concepts without hiding high-level abstractions. The same benefits apply to polymorphic numbers and strings.

@@

\paragraph{Advanced Type Feature Visualization}
The design of GeckoGraph enables the visualization of many advanced type-level features. \textbf{Kind visualization}: if the kind of type variables can be inferred, the kind information is consistently displayed in GeckoGraph. For example, in figure \ref{fig:advanced} (A), the variable \texttt{a} needs at least the kind \texttt{* -> *} because of its use on the right-hand side. GeckoGraph respects this kind information and displays it as a constructor type over an empty structure, indicated using a dotted outline.  \textbf{Qualified constraints}: GeckoGraph's type class notation naturally extends to support qualified constraints. In the type \texttt{forall b. A (a b) => a b}, GeckoGraph shows the scope type class requirement on \texttt{a b} (Fig. \ref{fig:advanced} B).
\textbf{Multiple Parameter Type Class}:  GeckoGraph supports multiple parameter type classes by using multiple shapes with the same color hue to indicate the different parameters of the same type class.  For example, for the type \texttt{A a b => a b},  GeckoGraph shows that the variables a and b both need an A class, but they are the different parameters of A (Fig. \ref{fig:advanced} C).

\paragraph{Precise Interactivity}
Modern programming environments often allow programmers to mouse over part of the source code to query detailed information, such as definition, references or documentation. However, with text-based source code, it is often hard to distinguish whether programmers want the most specific fragment under the cursor or larger blocks. Because of its graphical layout, GeckoGraph allows programmers to precisely select which part of a type signature they intend to query, that is, in Fig. \ref{fig:advanced} (D) when the user mouses over the type class box (orange square) under the second occurrence of \texttt{a} the type class it represents is revealed in detail.

@@

\paragraph{Language Agnostic}
GeckoGraph can be implemented in any language that uses static typing. In programming projects, GeckoGraph supports polyglot programming projects. Typical circumstances include projects using foreign function interfaces or multiple languages for client- and server-side programming. GeckoGraph provides a common notation to describe the functionality and features of systems. In teaching and learning programming languages, GeckoGraph removes the nomenclature difference in different programming languages.  For example, when describing algebraic data types, different language communities use various names: tuple, enum, struct, etc. It is important to realize that these are the same concepts and ignore the minute linguistic barriers.

@@

\section{Evaluation}
To evaluate the usefulness of GeckoGraph, we designed a controlled experiment in the form of a game called "Zero to Hero". The game contains 10 levels of varying difficulty. At each level, participants are asked to implement a function called "\texttt{zeroToHero}" using only a list of available functions. These available functions are different at each level, and the target types of Zero and Hero vary at each level. The details of each level are provided in the Appendix (Appendix \ref{levels}).

The experiment aims to study how polymorphic types are used and reasoned about during programming tasks. In particular, we studied how programmers scan and select potentially useful functions from a library and compare intended types and actual types during type errors.

\subsection{An example level}
We illustrate the task of the user study using level 4 of the game. At this level (Fig. \ref{fig:level-example}), the programmers are tasked with implementing the function \texttt{ zeroToHero :: Zero a b -> Hero b b}. The available functions are \texttt{f1::Zero a b -> Hero b a}, \texttt{f2::Zero a a -> Hero a a}, \texttt{f3::Zero a b -> Hero b a}, and \texttt{f4::Zero a b -> Hero b b}. Two generic functions \texttt{(\$)} and \texttt{(.)} are provided to improve the ergonomics of composing functions, but all tasks can be won without the use of generic functions. The possible solution and other details of the level can be found in the appendix (Appendix \ref{levels}).

@@

At each level of the game, the programmer must select the right functions to achieve the target result. In particular, for this level, participants must discover that only \texttt{f4} and \texttt{f2} are necessary to produce the desired results. An implement that satisfies the target type is \texttt{zeroTohero z = f2  (f4  z)}.

@@

\subsection{Hypothesis}
In programming tasks that involve reading and understanding polymorphic types, graphic notation using visual elements that provide high grouping strength (colors, shapes, sizes, and symbols) can improve the performance of such tasks compared to traditional text-based type notation. Our null hypothesis is that "Using graphic notation has no effect compared to traditional text-based type notation." This hypothesis and the task design were registered at the Open Science Foundation prior to data collection.

@@

Participants can skip a level during the game if they are stuck. We believe that it is normal for anyone to get stuck on a challenging task, and being stuck on one of the 10 tasks does not discount their qualitative input of the tool. We limit the number of skips that a participant can use during the game to four times so that submitting qualitative feedback without completing at least some levels is impossible.

@@

\begin{figure}[]
  \includegraphics[width=\linewidth]{figures/SuccessfulRate}
  \caption{\label{fig:success-rate}Success rate of each task with and without GeckoGraph, grouped by experience level. The figure is cropped from 70\% to 100\% for readability. In most tasks, GeckoGraph provides a small edge. However, significant differences were found in task 10 of the beginner group and task 10 of the familiar group. }
\end{figure}

@@

\begin{figure}[]
  \includegraphics[width=\linewidth]{figures/Qualitative}
  \caption{\label{fig:qualitative} The users rated scores of how intuitive (left), distracting (middle), and helpful (right) they found GeckoGraph. Overall, programmers consider GeckoGraph to be intuitive and not distracting. However, opinions are split on its helpfulness. }
\end{figure}

@@

\subsubsection{Low barrier to learn and understand}
GeckoGraph has some key similarities to traditional text-based type notation. GeckoGraph respects the left-to-right reading order. GeckoGraph uses the familiar symbolic name as the secondary encoding. GeckoGraph simulates the prefix notation in type constructors and the infix notation in type operators. With these considerations, we ensured that programmers were able to use GeckoGraph fluently with a minimal amount of training.

@@

\subsubsection{Color Encoding}
GeckoGraph highly relies on color hue as a main encoding. It provides a strong visual grouping  \cite{Zeng2023-jz}  for programmers to identify subtle patterns in types, such as the order and placement of substructures. However, the perception of color is different from person to person. This becomes an even bigger issue for color-blind or visually impaired programmers. Although GeckoGraph uses color-blind friendly schemes, it is only a method to avoid indistinguishable types and is not a strong guarantee of effectiveness. For this, we are exploring different encodings, such as patterns and shapes, to maximize the accessibility of GeckoGraph.

@@

\subsubsection{Pedagogical Applications}
We believe that GeckoGraph can be a valuable tool in teaching techniques and theories in programming languages that are difficult to convey in plain language. In fact, many participants in our study reported that they had no prior knowledge of Haskell programming and that they could understand the programming concepts in the game and complete all the puzzles with the help of GeckoGraph.

@@

% GeckoGraph, as depicted in this paper, uses a single color for each type variable and constant. In fact, to create enough visual contrast, the range of the hue spectrum may not provide enough values. In practice, GeckoGraph can encode type names using two or more colors, which will provide enough combinations.

>>>>

%% 
%% Copyright 2007-2024 Elsevier Ltd
%% 
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%% 
\documentclass[preprint,12pt]{elsarticle}

@@

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{lineno}
\usepackage[colorinlistoftodos,prependcaption,textsize=tiny]{todonotes}
\journal{Journal of Computer Languages}

@@

 \author{Shuai Fu}

@@

 \affiliation{organization={Monash University}, country={Australia}}

@@

@@

\linenumbers

@@

While type polymorphism is one of the oldest topics in programming language theory \cite{Cardelli1987-fp}, little research focuses on the \textit{usability} of polymorphic types. Hage argues that the expressiveness and power of type systems often come at the cost of usability~\cite{Hage2020-hg}. We aim to investigate the challenges of using polymorphic types and explore how to improve their usability with visualization and modern HCI techniques. To achieve this, we propose GeckoGraph, a graphical notation for types. GeckoGraph aims to complement traditional text-based type notation and make reading, understanding, and comparing types easier. GeckoGraph is prototyped and verified iteratively, leading to a design with visual clarity applicable to many programming contexts. Our study evaluating 714 participants' ability to solve a series of programming challenges using GeckoGraph versus text-based type annotation is, to our knowledge, the largest controlled user study of a functional programming tool or concept. The study results show that GeckoGraph helps improve programmers' ability to succeed in resolving the type challenges, especially for novice programmers.

@@

\subsection{Design of GeckoGraph}
The design of GeckoGraph focuses on visualizing types in functional languages (e.g., Haskell, ML). In this paper, we use Haskell as an example. As illustrated in this section, it can express basic types, polymorphic types, algebraic data types, and many advanced type-level features. However, GeckoGraph could also be used in imperative and multiparadigm languages such as TypeScript and Rust. We provide a GeckoGraph visualization tool avaiable online for multiple languages \todo{cite GeckoGraph website}.

We identified three main design goals for GeckoGraph based on the challenges of using polymorphic types~\cite{Jun2000-ec, Jun2000-yu} and how programmers tend to use type annotations to support their programming tasks~\cite{Justin_Lubin2021-yy}, as follows.

\paragraph{\textbf{(D1) Easy to learn}}\label{goal1} GeckoGraph should take little to no effort to learn. The rules to translate a text-based type notation to GeckoGraph should be minimal. Where possible, GeckoGraph should be intuitive to programmers who are familiar with text-based type notation.

@@

\paragraph{\textbf{(D3) Optimized for comparing and searching}} \label{goal3} GeckoGraph should aim to make the task of comparing two types easy, especially to make subtle differences in text-based notation harder to miss. This also includes the task of choosing an ideal function from a list of potential functions. For example, programmers search for a desired function from a documentation site with only partial knowledge of its type (e.g., the arity of a function, containing a known type, or type classes requirements).

\paragraph{Simple Types} 
Simple types, such as type variables and concrete types, are displayed in a cell \includegraphics[height=1em]{figures/SimpleType}: a solid-colored rectangular box with an angled corner on the top left. Each type identifier encodes a distinct color hue of the cell. Its first 1 or 2 letters are displayed inside the cell at the bottom left to provide familiarity (design goal \ref{goal1}) and a strong secondary encoding. The angled corner in the top left provides visual separation between two cells, even when the same color cells are next to each other, allowing GeckoGraph to be zoomed out to extremely small sizes (Section \ref{subsec:space}) without suffering readability (design goal \ref{goal2}).

@@

\begin{figure}[]
  \includegraphics[width=\linewidth]{figures/Indicator}
  \caption{
        \label{fig:indicator}
        An example of visual hirachy of the function indicator. The function indicator can be used to easily identify the arity of a function type by counting the connecting function indicators. For higher-order functions where functions are arguments or return values of other functions, it is very easy to see the ``level" of functions. 
  }
\end{figure}

\paragraph{Type Classes} 
Type classes are an intrinsic part of Haskell \cite{Hudak2007-kn}, and many other functional languages. In GeckoGraph, the type classes (e.g., \texttt{ (A a, B a) => a}) are indicated in the extended area below one or more GeckoGraph cells \includegraphics[height=1.2em]{figures/TypeClass.png}. Each type class required on a type variable displays as a square indicator aligned on the right of the extended area. In GeckoGraph, a type-class gliph is embeded in every type variable that requires this  class. This means when displaying the type \texttt{(==) :: Eq a => 
a -> a -> Bool} in GeckoGraph, the constraint \texttt{Eq} appears in twice on both occurrences of the type variable \texttt{a}.

The GeckoGraph type class's design promotes the type class placements rather than the type class names. Programmers can easily see where and how many type classes are required, but they may need an extra step (global legends of the color mapping or pop-up window) to identify the name of the type class. We believe that this is a well justified trade-off. The strength of this appraoch lies in situations when type class context is complex. When reading a type \texttt{(A a, A c, B a, B b, C b) => a -> b -> c}, programmers may need to switch back and forth to remember which type classes are needed on which variable. GeckoGraph uses approximity to help minimize the effort to associate each type variable with all its type classes (design goal \ref{goal2}).

\subsection{Strengths of GeckoGraph}\label{sec:benefits}

\paragraph{Generalization patterns in type classes}
A common source of confusion among novice Haskell programmers is distinguishing between functions that operate on generalized types and those that work with concrete types \cite{Tirronen2015-nr}. For instance, the type constraints {\tt Integral a => a} and {\tt Integer}, {\tt Floating a => a} and {\tt Float}, as well as {\tt Foldable} instances versus {\tt []} can be particularly challenging. Many Haskell learners may find the differences between concrete types and their generic counterparts irrelevant to their programming tasks, thereby complicating their quest for solutions. In the example shown in Fig. \ref{fig:example-foldable}, a programmer mistakenly supplies a \texttt{Char} instead of a \texttt{String} literal. The compiler responds with an error message that compares the provided \texttt{Char} type to a confusingly named type \texttt{t0 a0}. In this case, the information has been generalized too much and the programmer can hardly find their bearing to the original problem. This level of generalization can undermine a programmer's confidence in understanding the language and navigating type errors. In GeckoGraph, both list types and types with a Foldable instance share a similar structure. Therefore, the type \texttt{t0 a0} in the error message matches the shape of \texttt{[a]}. Bring programmer closer to the real issue, trying to match a complex type with a concrete simple type. This feature lowers the entry barrier for languages like Haskell, where even basic arithmetic functions {\tt +} and {\tt -} are defined on generic types. This approach lowers the barrier to introduce foundational language concepts without sacrificing generality.

@@

\paragraph{Advanced Type Feature Visualization}
The design of GeckoGraph enables the visualization of many advanced type-level features. \textbf{Kind visualization}: Kind system \cite{Jones1995-lm} is used in Haskell to indicate whether a type is type constructor (e.g. [], Maybe) or a type constant (e.g. Bool, Char), because of this, kind is often described as the type of types. Goanna infer the kind of type variables,  consistently display this information in GeckoGraph. For example, in figure \ref{fig:advanced} (A), the variable \texttt{a} is infered to have the kind \texttt{* -> *} because of its use on the right-hand side. GeckoGraph respects this kind information and displays the same kind level on the first occurance of \texttt{a}, but over an empty structure, indicated using a dotted outline. \textbf{Qualified constraints}: GeckoGraph's type class notation naturally extends to support qualified constraints. In the type \texttt{forall b. A (a b) => a b}, GeckoGraph shows the scope type class requirement on \texttt{a b} (Fig. \ref{fig:advanced} B).
\textbf{Multiple Parameter Type Class}:  GeckoGraph supports multiple parameter type classes by using multiple shapes with the same color hue to indicate the different parameters of the same type class. For example, for the type \texttt{A a b => a b},  GeckoGraph shows that the variables a and b both need an A class, but they are the different parameters of A (Fig. \ref{fig:advanced} C).

\paragraph{Precise Interactivity}
Modern programming environments often allow programmers to mouse over part of the source code to query detailed information, such as definition, references or documentation. However, with text-based source code, it is often hard to distinguish whether programmers want the most specific fragment under the cursor or larger blocks. Because of its graphical layout, GeckoGraph allows programmers to precisely select which part of a type signature they intend to query, that is, in Fig. \ref{fig:advanced} (D) when the user mouses over the type class box (orange square) under the second occurrence of \texttt{a} the type class it represents is revealed in detail, a sepecific query that is hard to hard to execute with point and click over text based type annotation.

@@

\paragraph{Language Agnostic}
GeckoGraph can be implemented in any language that exercises static typing. We identified applications of this feature in programming and education. In programming projects often use more than one programming languages, situations include projects using foreign function interfaces or multiple languages for client-side and server-side programming. In these cases, GeckoGraph provides a shared notation to describe the functionality and features of systems, dispide their text-based notations are different. In teaching and learning programming languages, GeckoGraph removes the nomenclature difference in different programming languages. For example, when describing the product type in algebraic data types, different language communities use various names: record, named tuple, struct, etc. GechoGraph helps to unify the same programming concept and ignore the minute nomenclature differences .

@@

\section{Evaluation}
To evaluate the usefulness of GeckoGraph, we designed and conducted a controlled experiment in the form of a game called "Zero to Hero". The game contains 10 levels of varying difficulty. At each level, participants are asked to implement a function named "\texttt{zeroToHero}" with the help of a limited list of provided functions. These functions are different at each level, and the target types of Zero and Hero vary at each level as well. The details of each level are provided in the Appendix (Appendix \ref{levels}).

The experiment aims to study how programmers use and reason about polymorphic type when performing programming tasks. In particular, we studied how programmers scan and select potentially useful functions from a library and compare intended types and actual types during type errors.

\subsection{An example level}
We illustrate the task of the user study using level 4 of the game. At this level (Fig. \ref{fig:level-example}), the programmers are tasked with implementing the function \texttt{ zeroToHero :: Zero a b -> Hero b b}. The available functions are \texttt{f1::Zero a b -> Hero b a}, \texttt{f2::Zero a a -> Hero a a}, \texttt{f3::Zero a b -> Hero b a}, and \texttt{f4::Zero a b -> Hero b b}. Two generic functions \texttt{(\$)} and \texttt{(.)} are provided to improve the ergonomics of composing functions, but all tasks can be completed without the use of generic functions. The possible solution and other details of the level can be found in the appendix (Appendix \ref{levels}).

@@

At each level of the game, the programmer must select the right set of functions, and compose them correctly to achieve the target result. Not all provided functions are useful. In particular, for this level, participants must discover that only \texttt{f4} and \texttt{f2} are necessary to produce the desired results. One implementation that satisfies the target type is \texttt{zeroTohero z = f2 (f4  z)}.

@@

\subsection{Hypothesis}
In programming tasks that involve reading and understanding polymorphic types, graphic notation using visual elements that provide higher grouping strength (colors, shapes, sizes, and symbols) can improve the performance of such tasks compared to traditional text-based type notation. Our null hypothesis is: {\it Using graphic notation has no effect compared to traditional text-based type notation.} This hypothesis and the task design were registered at the Open Science Foundation prior to data collection \todo{cite this}.

@@

Participants can skip a level during the game if they are stuck. We believe that it is normal for anyone to get stuck on a challenging task, and being stuck on one of the 10 tasks does not discount their feedback of the tool. We limit the number of skips that a participant can use during the game to four so that submitting qualitative feedback without completing at least some levels is impossible.

@@

The level 10 is the most cognatively taxing: the provided functions are very long and complex, and the solution is very challenging and deceptive, programmers have to utilize all the provided functions, and compese them eight time to reach the final position. Because level 10 require programmers to switch back and forth to remember what each function does. We speculate that this level require the mental agility of a Haskell expert to internalize these functions, and GeckoGraph help less experienced programmers achieve the similar level efficacy.

\begin{figure}[]
  \includegraphics[width=\linewidth]{figures/SuccessfulRate}
  \caption{\label{fig:success-rate}{\bf Success rate of each task with and without GeckoGraph, grouped by experience level.} The figure is cropped from 70\% to 100\% for readability. In most tasks, GeckoGraph provides a small edge. However, significant differences were found in task 10 of the beginner group and task 10 of the familiar group. }
\end{figure}

@@

It is interesting to note that programmers find GeckoGraph intuitive, and the more experienced they are, the more likely they find GeckoGraph easy to use. We believe this aligns with our design decision to make GeckoGraph resemble text-based type annotation. Those familiar with text-based type annotation will find GeckoGraph second nature to read. However, the correlation reverses when it comes to perceived helpfulness. The more experienced the users are, the less likely they are to find GeckoGraph helpful. We speculate this pattern occurs because users who are experts in text-based type annotations naturally prefer reading text-based notation and tend to ignore GeckoGraph, diminishing its usefulness. Conversely, novice users are drawn to GeckoGraph and find its visualization offers more clarity than text-based notations.

\begin{figure}[]
  \includegraphics[width=\linewidth]{figures/Qualitative}
  \caption{\label{fig:qualitative} {\bf The users rated scores of how intuitive (left), distracting (middle), and helpful (right) they found GeckoGraph.} Overall, programmers consider GeckoGraph to be intuitive and not distracting. However, opinions are split on its helpfulness. }
\end{figure}

@@

\subsubsection{Low barrier to learn and understand}
GeckoGraph has some key similarities to traditional text-based type notation. GeckoGraph respects the left-to-right reading order. GeckoGraph uses the familiar symbolic name as the secondary encoding. GeckoGraph simulates the prefix notation in type constructors and the infix notation in type operators. With these considerations, we ensured that programmers who were familiar with the Haskell language had no problem understanding and using GeckoGraph fluently with a minimal amount of training.

@@

\subsubsection{Color Encoding}
GeckoGraph highly relies on color hue as a main encoding. It provides a strong visual grouping  \cite{Zeng2023-jz}  for programmers to identify patterns in types, such as the order and placement of substructures. However, the perception of color is different from person to person. This becomes an even bigger issue for  visually impaired programmers. Although GeckoGraph uses color-blind friendly schemes, it is only a method to avoid indistinguishable types and is not a strong guarantee of effectiveness. For this, we are exploring different visual encoding mehods, such as patterns and shapes, to maximize the accessibility of GeckoGraph.

@@

\subsubsection{Pedagogical Applications}

We believe that GeckoGraph can be a valuable tool in teaching techniques and theories in programming languages that are difficult to convey in plain language. In fact, many participants in our study reported that they had no prior knowledge of Haskell programming and that they could understand the programming concepts in the game and complete all the puzzles with the help of GeckoGraph.

@@
